The Ky Fan Norms and Beyond: Dual Norms and Combinations for Matrix Optimization

### 2025-11-02 conf_article/neon.tex
edited a bit more and applied latex format

\section{\replaced[id = IK, comment={I think the old one was a bit misleading. Merged what previously had been 2 subsections into this one and improved the notation. Also removed the thing about trust region, it'll be referenced in a succeeding sections.}]{Update Step as Constrained Optimization}{Problem Statement}}

Training a neural network involves optimizing a function of weight matrices. This naturally leads us to consider the problem of minimizing a differentiable matrix function: 
\begin{equation}\label{eq:mat}
    F(\cdot)\colon \Rmn \to \R\,,\qquad F(\mX) \to \min_{\mX \in \Rmn}\,.
\end{equation}
We equip the matrix space $\Rmn$ with a standard dot product $\<\cdot, \cdot> \to \R$ and a norm $\norm{\cdot}\colon \Rmn \to \R_+$, which possibly does not coincide with the Frobenius norm $\normf{\cdot}$. The dual norm $\norm{\cdot}^\dagger\colon \Rmn \to \R_+$ associated with $\norm{\cdot}$ is defined as 
\begin{equation}\label{eq:dual_norm}
    \norm{\mG}^\dagger = \sup_{\mD \in \Rmn\,,\norm{\mD}\leq 1} \<\mG,\mD>\,.
\end{equation}

One approach to optimizing such functions is to employ an iterative algorithm based on Linear Minimization Oracle (LMO):
\begin{equation}\label{eq:lmo}
    \mathrm{LMO}(\mG) \in \argmin_{\mD \in \cS} \<\mG, \mD>\,,
\end{equation}
where $\mG$ is typically a gradient (possibly with momentum) of $F$ and $\cS \subset \Rmn$ is some set. The update rule used in Muon optimizer is uSCG \citep{pethick2025training}. In the most general case, which involves momentum and stochastic gradient $g(\mX^k, \xi^k)$, it can be written as
\begin{equation}\label{eq:usgd_update}
    \begin{aligned}
    \mM^{k} &= \alpha_{k} g(\mX^k, \xi^k) + (1 - \alpha_k)\mM^{k - 1}\,,\\
    \mX^{k + 1} &= \mX^k + \gamma_k\mathrm{LMO}(\mM^{k})\,.
    \end{aligned}
\end{equation}

Here we are particularly interested in the case when $\cS$ is a ball in the $\norm{\cdot}$ norm:
\begin{equation}\label{eq:ball}
    \cS = \cB_\eta = \{ \mD \in \Rmn \mid \norm{\mD} \leq \eta \}\,.
\end{equation}
In this case it can be easily shown that
\begin{equation} \label{eq:lmo_in_norm_ball}
    \argmin_{\mD \in \cS} \<\mG, \mD> = - \eta \{ \mD \in \cB_1 \mid \<\mG, \mD> = \norm{\mG}^\dagger\}\,.
\end{equation}
And update for $\mX$ in \cref{eq:usgd_update} becomes
\begin{equation}\label{eq:almost_trust_update}
    \mX^{k+1} = \mX^{k} - \gamma_k\eta \{\mD \in \cB_1 \mid \<\mM^k, \mD> = \norm{\mM^k}^\dagger\}\,.
\end{equation}
This formula will later allow us to easily compute updates for various norms.

\section{Different Norms $\norm{\cdot}$ imply different updates}

    Based on different norms $\norm{\cdot}$, we will simplify the update defined by the aforementioned equation:

    \begin{equation}\label{eq:our_update}
        \mX^{k+1} = \mX^{k} - \eta \{\Delta \in \cB_1 \mid \<\mM^k, \Delta> = \norm{\mM^k}^\dagger\}
    \end{equation}

    Throughout this work, we denote the singular value decomposition (SVD) of $\mM^k$ as $\mM^k = \mU \mSigma \mV^\top$, where $\mU = [u_1, u_2, \dots, u_r]$, $\mSigma = \diag(\sigma_1, \sigma_2, \dots, \sigma_r)$, and  $\mV = [v_1, v_2, \dots, v_r]$.

@@

Despite such similarities, no theory has been yet proposed that would reduce the matrix case to the optimization of a function of a singular values vector. If such a theory is developed, analysis of matrix algorithms like Muon will be greatly simplified.

\begin{table*}[t]
    \caption{lmo optimizers in Schatten $S_p$ norms and in $l_p$ norms. $g$ is the gradient. When it is a matrix, $g = \mU \mSigma \mV^\top$}
    \label{tbl:mat_vs_vec_lmo}
    \bgroup
    \def\arraystretch{1.2}
    \resizebox{\textwidth}{!}{
    \begin{tabular}{|l|c|c|c|}
    \hline
    Method & lmo constraint set $\mathcal D$ & lmo & Reference \\
    \hline
    \hline
    Normalized SGD & $l_2$-ball, $S_2$-ball & $-\eta \tfrac{g}{\norm{g}_2} = -\eta \tfrac{g}{\norm{g}_F}$ & \citep{hazan2015beyond} \\
    Momentum Normalized SGD & Ball in $l_2$, or Ball in $S_2$ & $-\eta \tfrac{g}{\norm{g}_2} = -\eta\tfrac{g}{\norm{g}_F}$ & \citep{cutkosky2020momentum}\\
    \hline
    SignSGD & Ball in Max-norm $l_\infty$ & $-\eta \sign(g)$ & \citep[Thm. 1]{bernstein2018signsgd} \\
    Signum  & Ball in Max-norm $l_\infty$ & $-\eta \sign(g)$ & \citep[Thm. 3]{bernstein2018signsgd} \\
    \hdashline
    Muon & Ball in Spectral $S_\infty$ & $-\eta UV^\top$ & \citep{jordan2024muon} \\
    \hline
    Gauss-Southwell Coordinate Descent & Ball in $l_1$ & $-\eta \{i: g_i \geq g_k \forall k\}$ & \citep[p. 19]{shi2016primer}\\
    \hdashline 
    Neon & Ball in Nuclear $S_1$ & $-\eta u_1 v_1^\top$ & This work\\
    \hline
    \end{tabular}
    }
    \egroup
    \end{table*}

@@

The practical drawback of the TRLAn is the absence of its implementation in PyTorch.
\begin{table}[h!]
    \centering
    \begin{tabular}{c|c|c|c}
    \hline
    Method                              & rtol        & k & time,s \\
    \hline 
    Power Iterations                    & 0.01        & 1 & 7.7\\ 
    SVDS (thick-restart Lanczos method) & 0.01        & 1 & 0.18\\
    PCA Low Rank (RSVD)                 & 0.01        & 1 & 1.15\\
    SVDS (thick-restart Lanczos method) & 0.01        & 10 & 0.47\\
    PCA Low Rank (RSVD)                 & 0.01        & 10 & 19.4\\
    SVDS (thick-restart Lanczos method) & 0.01        & 100 & 1.96\\
    PCA Low Rank (RSVD)                 & 0.01        & 100 & 170\\
    \end{tabular}
    \caption{k-rank updated comparison}
    Comparison of different numerical methods to calculate k-rank update on $5000\times5000$ matrix of real numbers, rtol is an error in Frobenius norm relative to the k-rank approximation. 
    \label{tbl:matrix_methods}
\end{table}
\section{Trust Region Bounds for L-smooth functions}
    First, we analyze the problem in the unstochastic case. From Corollary 1 of \citet{kovalev2025understanding}, we directly get the following result that matches lower bounds, as was noted in that article.

@@

    \begin{lemma}\label{lemma:stoch_tr}
    To reach the precision $\E{\min_{k=1\ldots K} \norm{\nabla f(\mX_k)}^\dagger} \leq \epsilon$ by ~\eqref{eq:our_update} under the assumptions \Cref{eq:variance}, \Cref{eq:L}, \Cref{eq:norm}, it is sufficient to choose the parameters as follows:
    \begin{align}
        \eta &= \cO\left(\min\left\{\frac{\epsilon}{L}, \frac{\epsilon^3}{\rho^2\sigma^2L}\right\}\right),
        \qquad
        \alpha = \cO\left(\min\left\{1, \frac{\epsilon^2}{\rho^2\sigma^2}\right\}\right),
        \\
        \label{eq:str_K_nonconvex}
        K &= \cO\left(\max\left\{
            \frac{\rho\sigma}{\epsilon},
            \frac{\rho^3\sigma^3}{\epsilon^3},
            \frac{L\Delta_0}{\epsilon^2},
            \frac{L\Delta_0\rho^2\sigma^2}{\epsilon^4}
        \right\}\right).
    \end{align}
    \end{lemma}

@@

\iffalse

\section{Idea}
\iffalse
The goal of the project is to make variations on Muon to speed it up. Recently, authors of \cite{bernstein2024oldoptimizernewnorm} have proposed to derive update step for an optimizer as the solution of the certain optimization problem. This approach can be utilized to derive Muon \cite{jordan2024muon}, a novel algorithm for fast training of neural networks. Instead of using spectral norm, used to derive Muon, we utilize nuclear norm to produce a new optimization algorithm.
\fi
\subsection{Problem (Project description)}

@@

    \item Neon (rank-1 version), Muon, AdamW and SGD are compared on MLP and CNN (see Figures~\ref{fig:mlp_epochs}, \ref{fig:cnn_epochs}, \ref{fig:mlp_time}, and~\ref{fig:cnn_time}). All methods work correctly, but again there is the problem with which one is the fastest (for now, it's SGD).
    \begin{figure}[h!]
        \center{\includegraphics[width=0.8\linewidth]{figs/mlp0605/loss_vs_epochs_mlp.pdf}}
        \caption{MLP: self.linear1 = nn.Linear(32*32*3, 512), self.linear2 = nn.Linear(512, 10), self.activ = nn.GELU()}
        \label{fig:mlp_epochs}
    \end{figure}
    \begin{figure}[h!]
        \center{\includegraphics[width=0.8\linewidth]{figs/mlp24/accuracy_vs_epochs_moderate.png}}
        \caption{CNN: 2 convolutional blocks, 2 fully connected layers, activation + dropout}
        \label{fig:cnn_epochs}
    \end{figure}
    \begin{figure}[h!]
        \center{\includegraphics[width=0.8\linewidth]{figs/mlp0605/accuracy_vs_time_mlp.pdf}}
        \caption{MLP: wallclock time measurements}
        \label{fig:mlp_time}
    \end{figure}
    \begin{figure}[h!]
        \center{\includegraphics[width=0.8\linewidth]{figs/mlp24/accuracy_vs_time_moderate.png}}
        \caption{CNN: wallclock time measurements}
        \label{fig:cnn_time}
    \end{figure}
    \begin{table}[h!]
        \centering
        \begin{tabular}{c|c|c|c}
        \hline
        Method                              & rtol        & k & time,s \\
        \hline 
        Power Iterations                    & 0.01        & 1 & 7.7\\ 
        SVDS (thick-restart Lanczos method) & 0.01        & 1 & 0.18\\
        PCA Low Rank (RSVD)                 & 0.01        & 1 & 1.15\\
        SVDS (thick-restart Lanczos method) & 0.01        & 10 & 0.47\\
        PCA Low Rank (RSVD)                 & 0.01        & 10 & 19.4\\
        SVDS (thick-restart Lanczos method) & 0.01        & 100 & 1.96\\
        PCA Low Rank (RSVD)                 & 0.01        & 100 & 170\\
        \end{tabular}
        \caption{k-rank updated comparison}
        Comparison of different numerical methods to calculate k-rank update on $5000\times5000$ matrix of real numbers, rtol is an error in Frobenius norm relative to the k-rank approximation of truncated svd. During the research it was noted that rsvd can give good and fast approximation for singular values, but the matrix of approximation is far from the one given by truncated svd, while Lanczos method gives good and fast approximation for a matrix, but not so good approximation for singular values. 
        \label{tab:matrix_methods_old}
    \end{table}
\end{enumerate}

>>>>

\section{\added[id = IK, comment={I think it goes well into one big section.}]{Update Step as Constrained Optimization}}

\subsection{\replaced[id = IK, comment={Merged what previously had been section with two subsections into this one and improved the notation. Also removed the thing about trust region, it'll be referenced in succeeding sections.}]{Linear Minimization Oracle}{Problem Statement}}

Training a neural network involves optimizing a function of weight matrices. For simplicity, we consider the problem of minimizing a differentiable function of a single matrix:
\begin{equation}\label{eq:mat}
	F(\cdot)\colon \Rmn \to \R\,,\qquad F(\mX) \to \min_{\mX \in \Rmn}\,.
\end{equation}
In practice, neural networks have multiple weight matrices $\mX_1, \mX_2, \ldots$, and the objective function depends on all of them. The framework we consider applies to each weight matrix independently, where the gradient $\mG_i = \nabla_{\mX_i} F$ for each matrix is computed via backpropagation, and the update rule is applied separately to each matrix.

We equip the matrix space $\Rmn$ with a standard dot product $\<\cdot, \cdot> \to \R$ and a norm $\norm{\cdot}\colon \Rmn \to \R_+$, which possibly does not coincide with the Frobenius norm $\normf{\cdot}$. The dual norm $\norm{\cdot}^\dagger\colon \Rmn \to \R_+$ associated with $\norm{\cdot}$ is defined as
\begin{equation}\label{eq:dual_norm}
	\norm{\mG}^\dagger = \sup_{\mD \in \Rmn\,,\norm{\mD}\leq 1} \<\mG,\mD>\,.
\end{equation}

One approach to optimizing such functions is to employ an iterative algorithm based on Linear Minimization Oracle (LMO):
\begin{equation}\label{eq:lmo}
	\mathrm{LMO}(\mG) \in \argmin_{\mD \in \cS} \<\mG, \mD>\,,
\end{equation}
where $\mG$ is typically a gradient (possibly with momentum) of $F$ and $\cS \subset \Rmn$ is some set. The update rule used in Muon optimizer is uSCG \citep{pethick2025training}. In the most general case, which involves momentum and stochastic gradient $g(\mX^k, \xi^k)$, it can be written as
\begin{equation}\label{eq:usgd_update}
	\begin{aligned}
		\mM^{k}     & = \alpha_{k} g(\mX^k, \xi^k) + (1 - \alpha_k)\mM^{k - 1}\,, \\
		\mX^{k + 1} & = \mX^k + \gamma_k\mathrm{LMO}(\mM^{k})\,.
	\end{aligned}
\end{equation}

Here we are particularly interested in the case when $\cS$ is a ball in the $\norm{\cdot}$ norm:
\begin{equation}\label{eq:ball}
	\cS = \cB_\eta = \{ \mD \in \Rmn \mid \norm{\mD} \leq \eta \}\,.
\end{equation}
In this case it can be easily shown that
\begin{equation} \label{eq:lmo_in_norm_ball}
	\argmin_{\mD \in \cS} \<\mG, \mD> = - \eta \{ \mD \in \cB_1 \mid \<\mG, \mD> = \norm{\mG}^\dagger\}\,.
\end{equation}
And update for $\mX$ in \cref{eq:usgd_update} becomes
\begin{equation}\label{eq:our_update}
	\mX^{k+1} = \mX^{k} - \gamma_k\eta \{\mD \in \cB_1 \mid \<\mM^k, \mD> = \norm{\mM^k}^\dagger\}\,.
\end{equation}
This formula will later allow us to easily compute updates for various norms.

\subsection{Different Norms $\norm{\cdot}$ imply different updates}

In this subsection we will derive and discuss updates induced by the choice of the norm in \cref{eq:our_update}. To describe those updates, the singular value decomposition (SVD) of $\mM^k$ is required. We denote SVD of $\mM^k$ as $\mM^k = \mU \mSigma \mV^\top$, where $\mU = [u_1, u_2, \dots, u_r]$, $\mSigma = \diag(\sigma_1, \sigma_2, \dots, \sigma_r)$, and  $\mV = [v_1, v_2, \dots, v_r]$.

@@

\deleted[id = IK, comment={This is essentially like saying uh oh the optimization in multy-dimensional space is so hard, why there is no algorithm to reduce this problem to 1-d case.}]{Despite such similarities, no theory has been yet proposed that would reduce the matrix case to the optimization of a function of a singular values vector. If such a theory is developed, analysis of matrix algorithms like Muon will be greatly simplified.}

\begin{table*}[t]
	\caption{lmo optimizers in Schatten $S_p$ norms and in $l_p$ norms. $g$ is the gradient. When it is a matrix, $g = \mU \mSigma \mV^\top$}
	\label{tbl:mat_vs_vec_lmo}
	\bgroup
	\def\arraystretch{1.2}
	\resizebox{\textwidth}{!}{
		\begin{tabular}{|l|c|c|c|}
			\hline
			Method                             & lmo constraint set $\mathcal D$ & lmo                                                         & Reference                            \\
			\hline
			\hline
			Normalized SGD                     & $l_2$-ball, $S_2$-ball          & $-\eta \tfrac{g}{\norm{g}_2} = -\eta \tfrac{g}{\norm{g}_F}$ & \citep{hazan2015beyond}              \\
			Momentum Normalized SGD            & Ball in $l_2$, or Ball in $S_2$ & $-\eta \tfrac{g}{\norm{g}_2} = -\eta\tfrac{g}{\norm{g}_F}$  & \citep{cutkosky2020momentum}         \\
			\hline
			SignSGD                            & Ball in Max-norm $l_\infty$     & $-\eta \sign(g)$                                            & \citep[Thm. 1]{bernstein2018signsgd} \\
			Signum                             & Ball in Max-norm $l_\infty$     & $-\eta \sign(g)$                                            & \citep[Thm. 3]{bernstein2018signsgd} \\
			\hdashline
			Muon                               & Ball in Spectral $S_\infty$     & $-\eta UV^\top$                                             & \citep{jordan2024muon}               \\
			\hline
			Gauss-Southwell Coordinate Descent & Ball in $l_1$                   & $-\eta \{i: g_i \geq g_k \forall k\}$                       & \citep[p. 19]{shi2016primer}         \\
			\hdashline
			Neon                               & Ball in Nuclear $S_1$           & $-\eta u_1 v_1^\top$                                        & This work                            \\
			\hline
		\end{tabular}
	}
	\egroup
\end{table*}

@@

The practical drawback of the TRLAn is the absence of its implementation in PyTorch.
\begin{table}[h!]
	\centering
	\begin{tabular}{c|c|c|c}
		\hline
		Method                              & rtol & k   & time,s \\
		\hline
		Power Iterations                    & 0.01 & 1   & 7.7    \\
		SVDS (thick-restart Lanczos method) & 0.01 & 1   & 0.18   \\
		PCA Low Rank (RSVD)                 & 0.01 & 1   & 1.15   \\
		SVDS (thick-restart Lanczos method) & 0.01 & 10  & 0.47   \\
		PCA Low Rank (RSVD)                 & 0.01 & 10  & 19.4   \\
		SVDS (thick-restart Lanczos method) & 0.01 & 100 & 1.96   \\
		PCA Low Rank (RSVD)                 & 0.01 & 100 & 170    \\
	\end{tabular}
	\caption{k-rank updated comparison}
	Comparison of different numerical methods to calculate k-rank update on $5000\times5000$ matrix of real numbers, rtol is an error in Frobenius norm relative to the k-rank approximation.
	\label{tbl:matrix_methods}
\end{table}
\section{Trust Region Bounds for L-smooth functions}
First, we analyze the problem in the unstochastic case. From Corollary 1 of \citet{kovalev2025understanding}, we directly get the following result that matches lower bounds, as was noted in that article.

@@

\begin{lemma}\label{lemma:stoch_tr}
	To reach the precision $\E{\min_{k=1\ldots K} \norm{\nabla f(\mX_k)}^\dagger} \leq \epsilon$ by ~\eqref{eq:our_update} under the assumptions \Cref{eq:variance}, \Cref{eq:L}, \Cref{eq:norm}, it is sufficient to choose the parameters as follows:
	\begin{align}
		\eta & = \cO\left(\min\left\{\frac{\epsilon}{L}, \frac{\epsilon^3}{\rho^2\sigma^2L}\right\}\right),
		\qquad
		\alpha = \cO\left(\min\left\{1, \frac{\epsilon^2}{\rho^2\sigma^2}\right\}\right),
		\\
		\label{eq:str_K_nonconvex}
		K    & = \cO\left(\max\left\{
		\frac{\rho\sigma}{\epsilon},
		\frac{\rho^3\sigma^3}{\epsilon^3},
		\frac{L\Delta_0}{\epsilon^2},
		\frac{L\Delta_0\rho^2\sigma^2}{\epsilon^4}
		\right\}\right).
	\end{align}
\end{lemma}

@@

\iffalse
	\section{Idea}
	\iffalse
		The goal of the project is to make variations on Muon to speed it up. Recently, authors of \cite{bernstein2024oldoptimizernewnorm} have proposed to derive update step for an optimizer as the solution of the certain optimization problem. This approach can be utilized to derive Muon \cite{jordan2024muon}, a novel algorithm for fast training of neural networks. Instead of using spectral norm, used to derive Muon, we utilize nuclear norm to produce a new optimization algorithm.
	\fi
	\subsection{Problem (Project description)}

@@

		\item Neon (rank-1 version), Muon, AdamW and SGD are compared on MLP and CNN (see Figures~\ref{fig:mlp_epochs}, \ref{fig:cnn_epochs}, \ref{fig:mlp_time}, and~\ref{fig:cnn_time}). All methods work correctly, but again there is the problem with which one is the fastest (for now, it's SGD).
		      \begin{figure}[h!]
			      \center{\includegraphics[width=0.8\linewidth]{figs/mlp0605/loss_vs_epochs_mlp.pdf}}
			      \caption{MLP: self.linear1 = nn.Linear(32*32*3, 512), self.linear2 = nn.Linear(512, 10), self.activ = nn.GELU()}
			      \label{fig:mlp_epochs}
		      \end{figure}
		      \begin{figure}[h!]
			      \center{\includegraphics[width=0.8\linewidth]{figs/mlp24/accuracy_vs_epochs_moderate.png}}
			      \caption{CNN: 2 convolutional blocks, 2 fully connected layers, activation + dropout}
			      \label{fig:cnn_epochs}
		      \end{figure}
		      \begin{figure}[h!]
			      \center{\includegraphics[width=0.8\linewidth]{figs/mlp0605/accuracy_vs_time_mlp.pdf}}
			      \caption{MLP: wallclock time measurements}
			      \label{fig:mlp_time}
		      \end{figure}
		      \begin{figure}[h!]
			      \center{\includegraphics[width=0.8\linewidth]{figs/mlp24/accuracy_vs_time_moderate.png}}
			      \caption{CNN: wallclock time measurements}
			      \label{fig:cnn_time}
		      \end{figure}
		      \begin{table}[h!]
			      \centering
			      \begin{tabular}{c|c|c|c}
				      \hline
				      Method                              & rtol & k   & time,s \\
				      \hline
				      Power Iterations                    & 0.01 & 1   & 7.7    \\
				      SVDS (thick-restart Lanczos method) & 0.01 & 1   & 0.18   \\
				      PCA Low Rank (RSVD)                 & 0.01 & 1   & 1.15   \\
				      SVDS (thick-restart Lanczos method) & 0.01 & 10  & 0.47   \\
				      PCA Low Rank (RSVD)                 & 0.01 & 10  & 19.4   \\
				      SVDS (thick-restart Lanczos method) & 0.01 & 100 & 1.96   \\
				      PCA Low Rank (RSVD)                 & 0.01 & 100 & 170    \\
			      \end{tabular}
			      \caption{k-rank updated comparison}
			      Comparison of different numerical methods to calculate k-rank update on $5000\times5000$ matrix of real numbers, rtol is an error in Frobenius norm relative to the k-rank approximation of truncated svd. During the research it was noted that rsvd can give good and fast approximation for singular values, but the matrix of approximation is far from the one given by truncated svd, while Lanczos method gives good and fast approximation for a matrix, but not so good approximation for singular values.
			      \label{tab:matrix_methods_old}
		      \end{table}
	\end{enumerate}
